### 2017-09-13 chapters/introduction/patient.tex
Add citation for validation of cardiac model
Update to latest version

\section{Personalization of Cardiac Biomechanics}
\label{sec:intro_personalization}
In this section we will go through the steps needed to personalize the
mechanical model developed in the previous section. We will start by
giving a brief overview of the interaction between medical imaging and
cardiac computational modeling. Then we will explain how the geometries
are extracted from the medical images and turned into a computational
meshes which we can use for finite element computations. Finally we
will explain how to incorporate additional data into model by means of
adjoint based data assimilation techniques.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/models.eps}
\caption{Interaction between medical imaging and computational
  models. Medical images are used to extract geometric
  structure. These geometries together with the computational models
  can later be validated against the medical images. Computational
  models can also be used together with the medical images to
  assimilate additional data in order to extraction clinically useful
  biomarkers. }
\label{fig:cardiac_imaging_model}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/raw.png}
    \caption{\label{fig:echopac_out_surf}}
  \end{subfigure}
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh.png}
    \caption{\label{fig:echopac_out_strain_mesh}}
  \end{subfigure}
  \begin{subfigure}[t]{0.38\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/bullseye/bullseye.pdf}
    \caption{\label{fig:bullseye_intro}}
  \end{subfigure}
\caption{Figure \ref{fig:echopac_out_surf} shows triangulated surfaces of the
      endo- and epicardium, and Figure
      \ref{fig:echopac_out_strain_mesh} shows structured mesh of the
      American Heart Association (AHA) segments. These surfaces are
      exported from the image segmentation tool EchoPac. Figure 
      \ref{fig:bullseye_intro} shows the American Heart Association (AHA)
      segments illustrated on a so called bullseye plot with name and
      numbers on the different segments. }
\label{fig:echopac_output}
\end{figure}

@@

As discussed in Section \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplished using a flat
base located at a prescribed location. In order to make the base flat
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the positive $x$
direction. To construct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
square fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left(ax_i + by_i + c  - z_i \right)^2.
\end{align}
Once the parameters for the basal plane is found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct cut size is found, the points above the plane is removed,
and the remaining surfaces are smoothed using the mesh generation tool GAMer
\cite{yu2008feature}.

@@

\begin{remark}
  The cavity volume is computed by generating the whole mesh, and the
  compute the volume using Equation \eqref{eq:volume_operator}. The cut size is
  found by using a one dimensional optimization algorithm with
  the objective functional representing the squared error between
  computed and measured volume.
\end{remark}

\subsubsection{Rule-based fiber architecture}
\label{sec:rule_based_fiber}
Both the passive and active properties of the myocardium depends on
the underlying muscle fiber architecture, and it is therefore of high
relevance to capture this architecture as accurately as
possible. Unfortunately, imaging technology does not currently provide a
way of extracting patients-specific fiber orientation \emph{in
  vivo}. Diffusion Tensor Imaging (DTI) can however be utilized to reconstruct
the fiber and laminar structure \emph{ex vivo}\cite{rohmer2007reconstruction}.

An alternative method for assigning myocardial fiber orientation is by
rule-based methods \cite{potse2006comparison, bayer2012novel}. 
In this thesis we have used an algorithm called the Laplace–Dirichlet
Rule-Based (LDRB) algorithm, proposed by Bayer et al \cite{bayer2012novel}.

@@

\begin{align}
  \begin{split}
    x &= a \sinh \mu \sin \nu \cos \theta  \\
    y &= a \sinh \mu \sin \nu \sin \theta  \\
    z &= a \cosh \mu \cos \nu,
  \end{split}
\end{align}
where $a$ is the focal point, $\nu \in [0, \pi]$ is the longitudinal
coordinate, $\theta \in [0, 2\pi]$ is the azimuth angle and $\mu$
is the radial coordinate. For an ellipse given by the equation
$\frac{x^2}{b^2} +\frac{y^2}{c^2}  = 1$ with $b$ being the semi-major
axis, and $c$ the semi-minor axis, the focal point is given by
\begin{align}
  a =\sqrt{b^2 - c^2}.
  \label{eq:focal_point}
\end{align}
The prolate spheroidal coordinate axes are illustrated in
Figure~\ref{fig:prolate_coord}. The inverse relation, going from the
Cartesian coordinate $(x,y,z)$ to the prolate spheroidal coordinate
$(\mu, \nu, \theta)$ given the focal point $a$, is a bit more
intricate, but can be computed as follows 
\begin{align}
  \begin{split}
    \nu &= \arccos (\tau) \\
    \mu &= \arccosh (\sigma) \\
    \theta &= \arctan \frac{z}{y} \\
    \tau &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2} \\
    & \quad + \sqrt{(x - a)^2 + y^2 + z^2} \big) \\
    \sigma &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2}\\
    & \quad - \sqrt{(x - a)^2 + y^2 + z^2}\big) 
  \end{split}
\end{align}
It is possible to estimate the focal point by taking $b$ in
\eqref{eq:focal_point} to be the maximum distance from the base to the
apex, and $c$ the maximum radius at the base. Using this estimate,
the base is assumed to be located approximately at the center of the
ellipsoid, i.e $\nu = \pi / 2$ \cite{land2015verification}.
% Compared to geometries used in
% benchmarking of cardiac mechanics problems, this is not a totally
% wrong assumption 
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate.png}
    \caption{\label{fig:prolate_coord}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate_cube.png}
    \caption{\label{fig:prolate_cube}}
  \end{subfigure}
\caption{Illustration of the prolate spheroidal coordinate system.}
\label{fig:prolate}
\end{figure}

@@

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
% The choice of reference geometry ..
%  choosing the geometry with the smallest
% volume as the reference geometry (i.e the end-systolic geometry)
% (references), the mid-diastolic or the end-diastolic ??

A general problem in the biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of and inflated version of the balloon. Having this picture in mind, it is
intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to estimate the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. We briefly
explain the steps in this methods.

@@

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some
medical image, which is loaded with a pressure $p_{\mathcal{I}}^{\mathrm{LV}}$ (assume LV only of
simplicity). We want to find the unloaded geometry 
$\Omega_{\mathcal{U}}$, so that when we load $\Omega_{\mathcal{U}}$ with $p_{\mathcal{I}}^{\mathrm{LV}}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{eq:intro_variational_form} after applying the pressure
$p_{\mathcal{I}}^{\mathrm{LV}}$. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To find
$\Xvec_{\mathcal{U}}$ we let 
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the applied pressure, do not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge, especially if the tissue is soft and the pressure in
the right ventricle is large. In this case the
right ventricular free wall might collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{chapters/introduction/figures/unloading_fail/unloading_fail.png}
\caption{Backward displacement method might fail for bi-ventricular
  geometries. This can be seen be noticing that the RV free wall is
  buckling and colliding with the septum. Here the image-based
  geometry is shown in black wire-frame, while the estimtated unloaded
geometry is shown in red.}
\label{fig:unloading_fail}
\end{figure}

An alternative to the backward displacement method that do not
necessary converge towards a fixed point but provides a more stable
way of obtaining an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}.
This algorithm is based on subtracting a factor $k$ times
the displacement obtained after inflating the image-based geometry to
the target pressure. To estimate the
unloaded configuration $\Xvec_{\mathcal{U}}$, we perform the following steps:
\begin{enumerate}
  \item Inflate $\Xvec_{\mathcal{I}}$ to target
    pressure $(p_{\mathcal{I}}^{\mathrm{LV}},
    p_{\mathcal{I}}^{\mathrm{RV}})$ to obtain
    $\mathbf{d} = \mathbf{d}(\Xvec_{\mathcal{I}})$.
  \item Given $k \in \mathbb{R}$, the unloaded configuration is given by
    \begin{align}
      \Xvec_{\mathcal{U}}(k) = \Xvec_{\mathcal{I}} - k \cdot \mathbf{d}(\Xvec_{\mathcal{I}}).
     \end{align}
     Further let $\mathbf{d}_k = \mathbf{d}(\Xvec_{\mathcal{U}}(k))$
     be the displacement obtained after inflating the unloaded
     configuration to the target pressure.
   \end{enumerate}
The uloading algorithm is based on finding $k$ so that the difference
between the image-based geometry and the inflated, unloaded geometry
is minimized, that is 
\begin{align}
  \underset{k}{\text{minimize}}
  \;\;  \mathrm{res} ( \Xvec_{\mathcal{I}}, \Xvec_{\mathcal{U}}(k) + \mathbf{d}_k)
  \label{eq:unloading_Raghavan}
\end{align}
with the residual given by the direct Hausdorff distance \cite{huttenlocher1993comparing}
\begin{align}
  \mathrm{res}(A, B) = \sup_{a \in A} \inf_{b \in B} d(a, b), 
\end{align}
and $d$ being the Euclidean distance.

@@

\subsubsection{Coupled material parameter estimation and unloading
  algorithm}
\label{sec:intro_coupled_material}
One thing to note is that the unloaded configuration depends upon the
chosen material parameters, which are generally also unknown. For
example, applying the backward displacement method to a geometry will
result in a smaller unloaded cavity volume if the material is softer.
Therefore, the material parameters estimation is coupled to the
problem of fined the unloaded, zero pressure geometry.
In \cite{nikou2016effects}, Nikou et al propose a method for estimating
material parameters as well the unloaded geometry by iterating between
the two methods, and terminate if the difference in unloaded cavity
volume is less than $5 \%$. This method is general and could be applied
using any unloading algorithm and any method to estimate the material
parameters.

Assume we are given an image-based LV geometry (which is not
end-diastole) with measured pressure $p_{\mathrm{LV}}$. Further assume
that we are given the pressure and  volume at end-diastole,
$p_{\mathrm{LV}}^{\mathrm{ED}}$ and $V_{\mathrm{LV}}^{\mathrm{ED}}$
respectively. Then we want to find
\begin{enumerate}
  \item \emph{the unloaded (zero pressure) geometry} so that when loaded with
    $p_{\mathrm{LV}}$ gives back the image based geometry, and
   \item \emph{material parameters}, so that when we load the
     estimated unloaded geometry with $p_{\mathrm{LV}}^{\mathrm{ED}}$
     it has a cavity volume equal to $V_{\mathrm{LV}}^{\mathrm{ED}}$.     
\end{enumerate}

The iterative strategy is presented in Algorithm \ref{alg:intro_unload_mat}.
\begin{algorithm}
  \caption{Uloaded material parameter estimation\label{alg:intro_unload_mat}}

\begin{algorithmic}[1]

  \State Initialize $\mathbf{m}_0$ \Comment{intial guess for material parameters}
  \State $i = 0$

  \While{difference in reference cavity volumes $>5 \%$ and $i < 15$}

  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent}{Find the
    unloaded geometry using any of the above methods and using
    $\mathbf{m}_{i}$ as material parameters.\strut} 
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent}{Estimate
    material parameters $\mathbf{m}_{i+1}$ so that when we load the
    estimated unloaded geometry with $p_{\mathrm{LV}}^{\mathrm{ED}}$
    it has a cavity volume equal to $V_{\mathrm{LV}}^{\mathrm{ED}}$. \strut}    
  \State $i = i + 1$
  \EndWhile

\end{algorithmic}
\end{algorithm}

Note that this method apply also to bi-ventricular geometries and to
cases when more measurement points are available for estimating the
material parameters.

@@

The other approach, called variational data assimilation, also known as
4D-var, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional. We will focus this approach since this is the approach
taken in this thesis. For a more thorough review of other
data assimilation techniques we refer to \cite{chapelle2013fundamental}.

@@

We will now explain the general setup and the main steps in variational data
assimilation. We will refer to
the \emph{state variable} as $\mathbf{w}$, which is our case
represents jointly the displacement and hydrostatic pressure
$\mathbf{w}=(\mathbf{u}, p)$.

The state variables are described by some \emph{physical forward
model} $\mathcal{M}$, which is our case is governed by the force balance
equation in \eqref{eq:force_balance_strong}. Moreover this model
typically depends on some parameters $\mu$ that we want to
adjust based on the observation at hand. We will refer to this
parameter as the \emph{control parameters}. For example, the control
parameters could be related to the stiffness of the myocardium which
we want to adjust to each individual patient based on some
clinical observations. We can therefore assume that the underlying
model can be written in the form $\mathcal{M}(\mathbf{w}, \mu) =
0$. Note that the state variable $\mathbf{w}$ typically depends on the
control parameters $\mu$, but we do not write this explicitly to
ease notation.

@@

\begin{remark}
  There exists gradient-free optimization methods that can be used to
  avoid computing the gradient. Examples include genetic algorithms or
  particle swarm algorithms. The drawback with such methods is that
  they typically requires a lot of functional evaluations. In our
  case this is very computationally expensive, because every functional
  evaluation requires a solution of the non-linear forward model. 
\end{remark}

@@

As discussed in Section \ref{sec:data_assimilation_pipeline}, a
solution to \eqref{eq:pde_unconstrained} satisfies the optimality
condition $D  \widehat{\mathcal{J}} = 0$. Here $D$ refers to the
total derivative with respect to the control $\mu$, while $D_{x}$
refers to the partial derivative with respect to $x$.

>>>>

\section{Personalization of Cardiac Biomechanics}
\label{sec:intro_personalization}
In this section we will go through the steps needed to personalize the
mechanical model developed in the previous section. We will start by
giving a brief overview of the interaction between medical imaging and
cardiac computational modeling. Then we will explain how the geometries
are extracted from the medical images and turned into computational
geometries which we can use for finite element computations. Finally we
will explain how to incorporate additional data into model by means of
adjoint based data assimilation techniques.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=\textwidth]{chapters/introduction/figures/models.eps}
\caption{Interaction between medical imaging and computational
  models. Medical images are used to extract geometric
  structure. These geometries together with the computational models
  can later be validated against the medical images. Computational
  models can also be used together with the medical images to
  assimilate additional data in order to extract clinically useful
  biomarkers. }
\label{fig:cardiac_imaging_model}
\end{figure}

@@

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/raw.png}
    \caption{\label{fig:echopac_out_surf}}
  \end{subfigure}
  \begin{subfigure}[t]{0.3\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh.png}
    \caption{\label{fig:echopac_out_strain_mesh}}
  \end{subfigure}
  \begin{subfigure}[t]{0.38\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/bullseye/bullseye.pdf}
    \caption{\label{fig:bullseye_intro}}
  \end{subfigure}
\caption{Figure \ref{fig:echopac_out_surf} shows triangulated surfaces of the
      endo- and epicardium, and Figure
      \ref{fig:echopac_out_strain_mesh} shows a structured mesh of the
      American Heart Association (AHA) segments. These surfaces are
      exported from the image segmentation tool EchoPac. Figure 
      \ref{fig:bullseye_intro} shows the American Heart Association (AHA)
      segments illustrated on a so called bullseye plot with name and
      numbers on the different segments. }
\label{fig:echopac_output}
\end{figure}

@@

As discussed in Section \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplish using a flat
base located at a prescribed location. In order to make the base flat,
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the positive $x$
direction. To construct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
square fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left(ax_i + by_i + c  - z_i \right)^2.
\end{align}
Once the parameters for the basal plane is found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct cut size is found, the points above the plane is removed,
and the remaining surfaces are smoothed using the mesh generation tool GAMer
\cite{yu2008feature}.

@@

\begin{remark}
  The cavity volume is computed by generating the whole mesh, and then
  computing the volume using Equation \eqref{eq:volume_operator}. The cut size is
  found by using a one-dimensional optimization algorithm with
  the objective functional representing the squared error between
  computed and measured volume.
\end{remark}

\subsubsection{Rule-based fiber architecture}
\label{sec:rule_based_fiber}
Both the passive and active properties of the myocardium depends on
the underlying muscle fiber architecture, and it is therefore of high
relevance to capture this architecture as accurately as
possible.

Unfortunately, imaging technology does not currently provide a
way of extracting patients-specific fiber orientation \emph{in
  vivo}. Diffusion Tensor MRI (DT-MRI) \cite{basser1994mr} can
be utilized to reconstruct the fiber and laminar structure \emph{ex
  vivo}\cite{rohmer2007reconstruction}, and recent studies have shown
promising results in measuring fiber orientations \emph{in vivo} using
DT-MRI \cite{toussaint2013vivo}. 
By using DT-MRI measurements of multiple hearts it is possible to build
statistical atlases \cite{peyrat2006towards} which could be used to
generate fiber fields based on transformation between a patient-specific
geometry and the geometries from the atlas \cite{vadakkumpadan2010modeling}.

An alternative method for assigning myocardial fiber orientation is by
rule-based methods \cite{potse2006comparison, bayer2012novel}. 
In this thesis we have used an algorithm called the Laplace – Dirichlet
Rule-Based (LDRB) algorithm, proposed by Bayer et al \cite{bayer2012novel}.

@@

\begin{align}
  \begin{split}
    x &= a \sinh \mu \sin \nu \cos \theta,  \\
    y &= a \sinh \mu \sin \nu \sin \theta,  \\
    z &= a \cosh \mu \cos \nu,
  \end{split}
\end{align}
where $a$ is the focal point, $\nu \in [0, \pi]$ is the longitudinal
coordinate, $\theta \in [0, 2\pi]$ is the azimuth angle and $\mu$
is the radial coordinate. For an ellipse given by the equation
$\frac{x^2}{b^2} +\frac{y^2}{c^2}  = 1$ with $b$ being the semi-major
axis, and $c$ the semi-minor axis, the focal point is given by
\begin{align}
  a =\sqrt{b^2 - c^2}.
  \label{eq:focal_point}
\end{align}
The prolate spheroidal coordinate axes are illustrated in
Figure~\ref{fig:prolate_coord}. The inverse relation, going from the
Cartesian coordinate $(x,y,z)$ to the prolate spheroidal coordinate
$(\mu, \nu, \theta)$ given the focal point $a$, is a bit more
intricate, but can be computed as follows 
\begin{align}
  \begin{split}
    \nu &= \arccos (\tau), \\
    \mu &= \arccosh (\sigma), \\
    \theta &= \arctan \frac{z}{y}, \\
    \tau &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2} \\
    & \quad + \sqrt{(x - a)^2 + y^2 + z^2} \big), \\
    \sigma &= \frac{1}{2a} \big( \sqrt{(x + a)^2 + y^2 + z^2}\\
    & \quad - \sqrt{(x - a)^2 + y^2 + z^2}\big). 
  \end{split}
\end{align}
It is possible to estimate the focal point by taking $b$ in
\eqref{eq:focal_point} to be the maximum distance from the base to the
apex, and $c$ the maximum radius at the base. Using this estimate,
the base is assumed to be located approximately at the center of the
ellipsoid, i.e $\nu = \pi / 2$ \cite{land2015verification}.
% Compared to geometries used in
% benchmarking of cardiac mechanics problems, this is not a totally
% wrong assumption 
\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate.png}
    \caption{\label{fig:prolate_coord}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/prolate_cube.png}
    \caption{\label{fig:prolate_cube}}
  \end{subfigure}
\caption{Illustration of the prolate spheroidal coordinate system.}
\label{fig:prolate}
\end{figure}

@@

\subsection{Choosing the reference geometry}
\label{sef:reference_geometry}
% The choice of reference geometry ..
%  choosing the geometry with the smallest
% volume as the reference geometry (i.e the end-systolic geometry)
% (references), the mid-diastolic or the end-diastolic ??
A general problem in biomechanics is that the geometry extracted from
imaging data is not the stress-free geometry, meaning the the geometry
that we observe is subjected to a physical load, e.g blood
pressure. In order for the stress and strain to be correct, we need to
find a geometry, so that when loaded with the measured loads, gives
back the geometry observed in the medical images. 
This is analogous to the problem of finding the geometry of a deflated balloon, given
that we have an image of and inflated version of the balloon. Having
this picture in mind, it is intuitive that such a geometry not necessarily is unique.
Several techniques have been applied in order to estimate the unloaded
geometry, such as the inverse design analysis (ID)
\cite{govindjee1996computational} or the modified updated Lagrangian
formulation (MULF) \cite{gee2010computational}. Other approaches
includes using the mid-diastolic or end-systolic geometry. Another popular
technique is a fixed point iteration scheme,  known as the
backward displacement method \cite{bols2013computational}. Figure
\ref{fig:unloading_lv} illustrate the backward displacement method applied
to an LV geometry. We briefly explain the steps in this methods.

@@

Suppose you are given geometry $\Omega_{\mathcal{I}}$ taken from some
medical image, which is loaded with a pressure $p_{\mathcal{I}}^{\mathrm{LV}}$ (assume LV only of
simplicity). We want to find the unloaded geometry 
$\Omega_{\mathcal{U}}$, so that when we load $\Omega_{\mathcal{U}}$ with $p_{\mathcal{I}}^{\mathrm{LV}}$,
you get back the starting geometry $\Omega_{\mathcal{I}}$. Let
$\Xvec_{\mathcal{I}}$ denote the coordinates in $\Omega_{\mathcal{I}}$, and
let $\mathbf{d}$ denote the displacement field which is the solution
of \eqref{eq:intro_variational_form} after applying the pressure
$p_{\mathcal{I}}^{\mathrm{LV}}$. What we want, is to find
$\Xvec_{\mathcal{U}}$ so that $\Xvec_{\mathcal{I}} =
\Xvec_{\mathcal{U}} + \mathbf{d}(\Xvec_{\mathcal{U}})$. To find
$\Xvec_{\mathcal{U}}$ we let 
\begin{align}
  \Xvec_{n+1} = g(\Xvec_n) =  \Xvec_{\mathcal{I}} - \mathbf{d}(\Xvec_n),
  && \Xvec_0 = \Xvec_{\mathcal{I}}.
\end{align}
According to Banach Fixed Point Theorem, the mapping $g$ has a
fixed-point if $\| \nabla g \|_{\infty} = \| \nabla \mathbf{d}
\|_{\infty} < 1$. This means that as long as the deformation
resulting from the applied pressure, do not result in a stretch that
is more than $100 \%$, this method will converge. Of course the
question about convergence of the fixed point method is one
thing. Another thing is the convergence of the nonlinear solver.
For bi-ventricular geometries, the backward displacement method might
fail to converge, especially if the tissue is soft and the pressure in
the right ventricle is high. In this case the
right ventricular free wall might collided with the septal wall when
subtracting the displacement. An example of this is illustrated in
Figure~\ref{fig:unloading_fail}.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\textwidth]{chapters/introduction/figures/unloading_fail/unloading_fail.png}
\caption{Backward displacement method might fail for bi-ventricular
  geometries. This can be seen be noticing that the RV free wall is
  buckling and colliding with the septum. Here the image-based
  geometry is shown in black wire-frame, while the estimated unloaded
geometry is shown in red.}
\label{fig:unloading_fail}
\end{figure}

An alternative to the backward displacement method that do not
necessary converge towards a fixed point but provides a more stable
way of obtaining an unloaded configuration was proposed in Ragahavan
et. al \cite{raghavan2006non}. In this algorithm, only one iteration of the
backward displacement method is applied but the subtracted
displacement is multiplied by a scalar $k$, which could be estimated
to minimizing some residual.
% \begin{align}
%   \Xvec_{\mathcal{U}}(k) = \Xvec_{\mathcal{I}} - k \cdot \mathbf{d}(\Xvec_{\mathcal{I}}).
% \end{align}

% This algorithm is based on subtracting a factor $k$ times
% the displacement obtained after inflating the image-based geometry to
% the target pressure. To estimate the
% unloaded configuration $\Xvec_{\mathcal{U}}$, we perform the following steps:
% \begin{enumerate}
%   \item Inflate $\Xvec_{\mathcal{I}}$ to target
%     pressure $(p_{\mathcal{I}}^{\mathrm{LV}},
%     p_{\mathcal{I}}^{\mathrm{RV}})$ to obtain
%     $\mathbf{d} = \mathbf{d}(\Xvec_{\mathcal{I}})$.
%   \item Given $k \in \mathbb{R}$, the unloaded configuration is given by
%     \begin{align}
%       \Xvec_{\mathcal{U}}(k) = \Xvec_{\mathcal{I}} - k \cdot \mathbf{d}(\Xvec_{\mathcal{I}}).
%      \end{align}
%      Further let $\mathbf{d}_k = \mathbf{d}(\Xvec_{\mathcal{U}}(k))$
%      be the displacement obtained after inflating the unloaded
%      configuration to the target pressure.
%    \end{enumerate}
% The uloading algorithm is based on finding $k$ so that the difference
% between the image-based geometry and the inflated, unloaded geometry
% is minimized, that is 
% \begin{align}
%   \underset{k}{\text{minimize}}
%   \;\;  \mathrm{res} ( \Xvec_{\mathcal{I}}, \Xvec_{\mathcal{U}}(k) + \mathbf{d}_k)
%   \label{eq:unloading_Raghavan}
% \end{align}
% with the residual given by the direct Hausdorff distance \cite{huttenlocher1993comparing}
% \begin{align}
%   \mathrm{res}(A, B) = \sup_{a \in A} \inf_{b \in B} d(a, b), 
% \end{align}
% and $d$ being the Euclidean distance.

@@

\subsubsection{Coupled material parameter estimation and unloading
  algorithm}
\label{sec:intro_coupled_material}
One thing to note is that the unloaded configuration depends upon the
chosen material parameters, which are generally also unknown. For
example, applying the backward displacement method to a geometry will
result in a smaller unloaded cavity volume if the material is softer.
Therefore, the material parameters estimation is coupled to the
problem of finding the unloaded, zero pressure geometry.
In Reference \cite{nikou2016effects}, Nikou et al propose a method for estimating
material parameters and the unloaded geometry by iterating between
the two methods, and terminating if the difference in unloaded cavity
volume is less than $5 \%$. This method is general and could be applied
using any unloading algorithm and any method to estimate the material
parameters. In the last paper in this thesis, we analyze this method
and shows that it actually converges to the correct solution are are
stable to noise in the measurements.

% Assume we are given an image-based LV geometry (which is not
% end-diastole) with measured pressure $p_{\mathrm{LV}}$. Further assume
% that we are given the pressure and  volume at end-diastole,
% $p_{\mathrm{LV}}^{\mathrm{ED}}$ and $V_{\mathrm{LV}}^{\mathrm{ED}}$
% respectively. Then we want to find
% \begin{enumerate}
%   \item \emph{the unloaded (zero pressure) geometry} so that when loaded with
%     $p_{\mathrm{LV}}$ gives back the image based geometry, and
%    \item \emph{material parameters}, so that when we load the
%      estimated unloaded geometry with $p_{\mathrm{LV}}^{\mathrm{ED}}$
%      it has a cavity volume equal to $V_{\mathrm{LV}}^{\mathrm{ED}}$.     
% \end{enumerate}

% The iterative strategy is presented in Algorithm \ref{alg:intro_unload_mat}.
% \begin{algorithm}
%   \caption{Uloaded material parameter estimation\label{alg:intro_unload_mat}}

% \begin{algorithmic}[1]

%   \State Initialize $\mathbf{m}_0$ \Comment{intial guess for material parameters}
%   \State $i = 0$

%   \While{difference in reference cavity volumes $>5 \%$ and $i < 15$}

%   \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent}{Find the
%     unloaded geometry using any of the above methods and using
%     $\mathbf{m}_{i}$ as material parameters.\strut} 
%   \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent}{Estimate
%     material parameters $\mathbf{m}_{i+1}$ so that when we load the
%     estimated unloaded geometry with $p_{\mathrm{LV}}^{\mathrm{ED}}$
%     it has a cavity volume equal to $V_{\mathrm{LV}}^{\mathrm{ED}}$. \strut}    
%   \State $i = i + 1$
%   \EndWhile

% \end{algorithmic}
% \end{algorithm}

% Note that this method apply also to bi-ventricular geometries and to
% cases when more measurement points are available for estimating the
% material parameters.

@@

The other approach, called variational data assimilation, also known as
4D-var, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional. We will focus this approach since this is the approach
taken in this thesis. For a more thorough review of other
data assimilation techniques we refer to the paper by Chapelle
et. al \cite{chapelle2013fundamental}.

@@

We will now explain the general setup and the main steps in variational data
assimilation, and will denote the \emph{state variable} by
$\mathbf{w}$, which is our case represents jointly the displacement
and hydrostatic pressure $\mathbf{w}=(\mathbf{u}, p)$.

The state variables are described by some \emph{physical forward
model} $\mathcal{M}$, which is our case is governed by the force balance
equation in \eqref{eq:force_balance_strong}. Moreover this model
typically depends on some parameters $\mu$ that we want to
adjust based on the observations at hand. We will refer to these
parameters as the \emph{control parameters}. For example, the control
parameters could be related to the stiffness of the myocardium which
we want to adjust to each individual patient based on some
clinical observations. We can therefore assume that the underlying
model can be written in the form $\mathcal{M}(\mathbf{w}, \mu) =
0$. Note that the state variable $\mathbf{w}$ typically depends on the
control parameters $\mu$, but to ease notation, we do not write this
dependency explicitly

@@

\begin{remark}
  There exists gradient-free optimization methods that can be used to
  avoid gradient computations. Examples include genetic algorithms or
  particle swarm algorithms. The drawback with such methods is that
  they typically requires a lot of functional evaluations. In our
  case functional evaluations are very computationally expensive,
  because every functional evaluation requires a solution of the
  non-linear forward model.  
\end{remark}

@@

Note that a solution to \eqref{eq:pde_unconstrained} satisfies the optimality
condition $D  \widehat{\mathcal{J}} = 0$. In the following, $D$ refers
to the total derivative with respect to the control $\mu$, while
$D_{x}$ refers to the partial derivative with respect to $x$.
